%! TEX root = master.tex
\lecture{19}{25/11}

\begin{theorem}[]
	Let $X$ and $Y$ be compact topological spaces.
	Then $X \times Y$ is compact.
\end{theorem}

\begin{proof}
	Let
	\[
		\mathcal U = \left\{
			U_i \times V_i: i \in I
		\right\}
	\]
	be an open cover of $X \times Y$ by basis sets.
	Let $p \in X$ and set
	\[
		\mathcal U^p = \left\{
			U \times V: U \times V \in \mathcal U, p \in U
		\right\}.
	\]
	Observe that
	$
		\left\{
			p
		\right\}
		\times Y
		\subset
		\bigcup_{U \times V \in \mathcal U^p} U \times V
	$
	and so
	$
		Y =
		\bigcup_{U \times V \in \mathcal U^p} V
	$
	is an open cover of $Y$.
	As $Y$ is compact, there is a finite set
	$\mathcal U_f^p \subset \mathcal U^p$
	such that
	$
		\left\{
			p
		\right\}
		\times Y 
		\subset
		\bigcup_{U \times V \in \mathcal U_f^p} U \times V.
	$
	Hence
	$
		p \in \bigcap_{U \times V \in \mathcal U_f^p} U = U_p
	$
	is an open neighbourhood of $p$.
	Thus $X = \bigcup_{p \in X} U_p$
	is an open cover.
	As $X$ is compact, there is $p_1, \ldots, p_n \in X$
	such that $X = \bigcup_{i=1}^n U_{p_i}$.
	Note
	\[
		U_{p_i} \times Y \subset
		\bigcup_{U \times V \in \mathcal U_f^p} U \times V
	\]
	and so
	\[
		X \times Y =
		\bigcup_{i=1}^n \left( 
			\bigcup_{U \times V \in \mathcal U_f^p} U \times V 
		\right). \qedhere
	\]
\end{proof}

\begin{theorem}[Heine-Borel]
	A subset of $\R^n$ is compact if and only if it is closed and bounded.
\end{theorem}

\begin{proof}
	$(\implies)$: let $A \subset \R^n$ be compact.
	As $\R^n$ is Hausdorff, $A$ is closed.
	Observe that
	\[
		A = \bigcup_{n=1}^\infty \left( 
			A \cap B(\bm 0; n) 
		\right)
	\]
	is an open cover of $A$.
	Hence there is $n_1, \ldots, n_r \in \N$ with $n_1 < \ldots < n_r$
	such that
	\[
		A =
		\bigcup_{n=1}^r \left( 
			A \cap B(\bm n; n) 
		\right)
		= A \cap B(\bm 0; r).
	\]
	Hence $A \subset B(\bm 0; r)$ is bounded.

	$(\impliedby)$: let $A \subset \R^n$ be closed and bounded.
	Let $R . )$ such that $A \subset [-R, R]^n$.
	Hence $A$ is a closed subset of a compact space, so itself is compact.
\end{proof}

\begin{examples}
	\begin{enumerate}
		\item
			We claim that $S^n$ is compact.
			Let
			\[
				f: S^n \to \R, \qquad
				f(x_1, \ldots, x_{n+1})
				= \sum^{n+1}_{i=1} x_i^2.
			\]
			$f$ is clearly continuous and $\{1\}$ is closed.
			Hence
			$
				f^{-1}\left( 
					\left\{
						1
					\right\} 
				\right) = S^n
			$
			is closed.
			Clearly $S^n$ is bounded and $S^n \subset \R^{n+1}$;
			hence, $S^n$ is compact.

		\item
			$T^n = S^1 \times S^1$ is clearly compact.

		\item
			We claim
			$
				X = \left\{
					\bm x \in \R^3:
					x_1^2 + x_2^2 - x_3^3 = 1
				\right\}
			$
			is not compact.
			Indeed, consider the sequence
			\[
				\left\{
					\left( 
						n,0,\left( 
							n^2 - 1 
						\right)^{\frac13} 
					\right)
				\right\}_{n \in \N}
				\subset X.
			\]
			But the first coordinate of this sequence is unbounded;
			hence $X$ is unbounded.
			Thus, as $X \subset \R^3$, it is not compact.
	\end{enumerate}
\end{examples}

\begin{corollary}
	Let $X$ be a compact space and $f: X \to \R$ be a continuous function.
	Then $f$ attains its minimum and maximum.
\end{corollary}

\begin{proof}
	$f(X) \subset \R$ is compact, and so it is closed and bounded.
	Thus $\sup f$ and $\inf f$ exist and are limit points of $f(X)$.
	But as $f$ is closed, it contains all its limit points.
	Hence $f$ attains its minimum and maximum.
\end{proof}

\begin{theorem}[Bolzano-Weierstrass]
	An infinite subset of a compact space must have a limit point.
\end{theorem}

\begin{proof}
	Suppose not.
	Let $A \subset X$ be an infinite subset of a compact space with no limit
	points.
	So for every $p \in X$ there is $U_p \subset X$ such that
	$
	\left( 
		U_p \setminus \left\{
			p
		\right\} 
	\right)
	\cap A = \varnothing.
	$
	Now
	$
		\left\{
			U_p
		\right\}_{p \in X}
	$
	is an open cover of $X$.
	So there is $p_1, \ldots, p_n$ suich that
	$
		X = \bigcup_{i=1}^n U_{p_i}.
	$
	But
	\[
		A
		= A \cap X
		= \bigcup_{i=1}^n (A \cap U_{p_i})
		\subset \left\{
			p_1, \ldots, p_n
		\right\};
	\]
	a contradiction.
\end{proof}

