%! TEX root = master.tex
\lecture{16}{12/11}

\begin{proposition}[]
	Let $X$ be a space,
	$Z \subset X$ be connected, and $Y \subset X$ with 
	$Z \subset Y \subset \overline Z$.
	Then $Y$ is connected.
	In particular, the closure of a connected set is connected.
\end{proposition}

\begin{proof}
	Suppose $Y = A \cup A'$ where $A$ and $A'$ are open and disjoint.
	So there is open $U,U' \subset X$ such that $A = U \cap Y$ and
	$A' = U' \cap Y$.
	As $Z \subset Y$,
	\begin{align*}
		Z &= Y \cap Z \\
		  &= (A \cup A') \cap Z \\
		  &= (A \cap Z) \cup (A' \cap Z) \\
		  &= (U \cap Z) \cup (U' \cap Z).
	\end{align*}
	Now observe
	\[
		(U \cap Z) \cap (U; \cap Z) \subset (U \cap Y) \cap (U' \cap Y)
		= varnothing
	\]
	and so $U \cap Z$ and $U' \cap Z$ are disjoint clopen sets in $\Z$.
	$\Z$ is clopen so without loss of generality assume that
	$Z = U \cap Z$ and $\varnothing = U' \cap Z$.
	So $Z \subset X \setminus U'$, a closed set.
	So $Y \subset \overline \Z \subset X \setminus U'$.
	Thus $Y \cap A' = A' = \varnothing$
	and $Y \cap A = A = Y$.
	Hence $Y$ is connected.
\end{proof}

\begin{corollary}
	Let $X$ be a topological space and $Z \subset X$.
	If $Z$ is connected and dense in $X$, then $X$ is connected.
\end{corollary}

\begin{proposition}[]
	Let $\mathcal A = \left\{
		A_i: i \in I
	\right\}$
	be a collection of subsets of a topological space $X$ such that
	$\bigcup_{i \in I} A_i = X$.
	Assume that $A_i$ is connected and for each $i,j \in I$ we hav
	$A_i \cap A_j \neq \varnothing$.
	Then $X$ is connected.
\end{proposition}

\begin{proof}
	Suppose $U$ and $V$ are open subsets of $X$ and 
	$U \cup V  = X$ and $V \cap V = \varnothing$.
	Fix $\varepsilon \in I$.
	Then $U \cap A_\varepsilon$ and $V \cap A_\varepsilon$ are
	complementary open subsets of $A_\varepsilon$.
	Without loss of generality, assume that $U \cap A_\varepsilon
	= A_\varepsilon$ and $V \cap A_\varepsilon = \varnothing$.
	If $i \in I$, $U \cap A_i$ and $V \cap A_i$ are complementary
	clopen sets of $A_i$.
	Noting that
	\[
		(U \cap A_i) \cap A_\varepsilon
		= (U \cap A_\varepsilon) \cap A_i
		= A_\varepsilon \cap A_i \neq \varnothing.
	\]
	Hence $U \cap A_i \neq \varnothing$.
	Hence $U \cap A_i = A_i$.
	Thus $A_i \subset U$ for every $i \in I$.
	Hence
	$X = \bigcup_{i\in I} A_i \subset U \subset X$
	and so $X = U$ and $V = \varnothing$.
\end{proof}

\begin{theorem}[]
	If $X$ and $Y$ are connected,
	then $X \times Y$ is connected.
\end{theorem}

\begin{proof}
	Pick $p \in X$ and $q \in Y$.
	Now $\left\{
		p
	\right\} \times Y \cong Y$ and $X \times \left\{
		q
	\right\} \cong X$ so both are connceted.
	As $(p,q)$ are in both sets,
	$(X \times \left\{
		q
	\right\}) \cup (\left\{
		p
	\right\} \times Y)$ is also connected.
	Now
	\[
		X \times Y = \bigcup_{q \in Y} \left( 
			\left( 
				X \times \left\{
					q
				\right\} 
			\right) \cup \left( 
				\left\{
					p
				\right\} \times Y 
			\right) 
		\right)
	\]
	and by the previous proposition, $X \times Y$ is connected.
\end{proof}

\begin{example}[]
	$\R^n$ is thus connected.
	$B^n$ is connected since $B^n \cong \R^n$
	and similarly $D^n = \overline{B^n}$ is also connected.
	Finally $T^n$ is the product of $n$ connected spaces,
	and thus it is connected.
\end{example}

\begin{example}
	Now consider $S^n$.
	We can define $S^n = U^n \cup L^n$ where
	\begin{align*}
		U^n &= S^n \cap \left\{
			\left( 
				x_1, \ldots, x_{n+1} 
			\right): x_1 \geq 0
		\right\} \\
		L^n &= S^n \cap \left\{
			\left( 
				x_1, \ldots, x_{n+1}: x_1 \leq 0 
			\right)
		\right\}.
	\end{align*}
	Now we define $f: D^n \to U^n$ by
	\[
		(x, \ldots, x_n) \mapsto \left(
			\sqrt{1 - x_2^2 - \ldots - x_{n+1}^2}, x_2, \ldots, x_{n+1}
		\right)
	\]
	and $f^{-1} : U^n \to D^n$ by
	\[
		(x_1, \ldots, x_{n+1}) \mapsto
		(x_2, \ldots, x_{n+1})
	\]
	and so $U^n \cong D^n$.
	Simiarly, $L^n \cong D^n$ and as
	$U^n \cap L^n \neq \varnothing$, $S^n$ is connected.
\end{example}

\begin{definition}[Component]
	A \emph{component} of a topological space $X$ is a maximal connected
	space.
\end{definition}

Note that each point is in a unique component.
Otherwise, if $p \in C_1, C_2$ for distinct components $C_1$
and $C_2$, then $C_1 \cap C_2 \neq \varnothing$ and
$C_1 \cup C_2$ is connected.
Then either $C_1$ is not maximal or $C_2$ is not maximal.

\begin{proposition}[]
	Each component of a topological space $X$ is closed.
\end{proposition}

\begin{proof}
	Let $C$ be a compoment of $X$.
	As $C$ is connected, so is $\overline C$.
	As $C$ is maximal and $C \subset \overline C$,
	$C = \overline C$.
\end{proof}

\begin{examples}
	\begin{enumerate}
		\item
		Let $X$ be a connected topological space.
		Then $X$ is the only component of itself.

		\item
		Let $X$ be a discrete topological space.
		Then every singleton set is a component since any set with more than one
		point is not connected.

		\item
		In $\Q$, the components are singleton sets.
		Suppose $x, y \in A \subset \Q$ such that $x < y$.
		Choose an irrational $\alpha \in (x,y)$.
		Then 
		$
			A = \left( 
				(-\infty, \alpha) \cap A 
			\right) \cup \left( 
				(\alpha, \infty) \cap A %s/textbf/emph/g
			\right)
		$.
	\end{enumerate}
\end{examples}
