%! TEX root = master.tex
\lecture{14}{6/11}

\begin{proposition}[]
	Let $X$ and $Y$ be spaces, and $X \times Y$ given the product topology.
	Then the projection maps
	\[
		\pi_X: X \times Y \to X, \qquad
		\pi_Y: X \times Y \to Y
	\]
	are continuous and they map open sets to open sets.
	Furthermore, the product topology is the smallest topology such that both
	projection maps are continuous.
\end{proposition}

\begin{proof}
	Note that if $U \subset X$ is open then
	$\pi^{-1}_X(U) 
	= U \times Y \in \mathcal B_{X \times Y} 
	\subset \tau_{X \times Y}$.
	Hence $\pi_X$ is continuous, and similarly so is
	$\pi_Y$.
	
	Now let $\tau$ be a topology on $X \times Y$ such that
	$\pi_X$ and $\pi_Y$ are continuous with respect to $\tau$.
	Let $U \subset X$ and $V \subset Y$ be open. 
	Then
	\[
		U \times V
		= (U \times Y) \cap (X \times V)
		= \pi_X^{-1} (U) \cap \pi_Y^{-1}(V)
		\in \tau
	\]
	and so $\mathcal B_{X \times Y} \subset \tau$.
	So arbitrary unions of elements of $\mathcal B_{X \times Y}$
	are elements of $\tau$ and so $\tau_{X \times Y} \subset \tau$.
\end{proof}

\begin{proposition}[]
	Let $X$, $Y$, and $Z$ be topological spaces.
	A function $f: Z \to X \times Y$ is continuous
	if and only if
	$\pi_x \circ f$ and $\pi_y \circ f$ are continuous.
\end{proposition}

\begin{proof}
	$(\implies)$: composition of continuous functions are continuous.
	$(\impliedby)$: let $U \times V \in \mathcal B_{X \times Y}$.
	Then
	\begin{align*}
		f^{-1}(U \times V)
		&= f^{-1}\left( (U \times Y) \cap (X \times V) \right) \\
		&= f^{-1}(U \times Y) \cap f^{-1}(X \times V) \\
		&= f^{-1}(\pi_x^{-1}(U)) \cap f^{-1}(\pi_Y^{-1}(X \times V)) \\
		&= (\pi_x \circ f)^{-1}(U) \cap (\pi_y \circ f)^{-1}(V)
	\end{align*}
	and, by Theorem \ref{the:top-spaces-equiv}, $f^{-1}(U \times V)$
	is open.
	Thus $f$ is continuous.
\end{proof}

\begin{example}[]
	Let $X$ be a space and $f: X \to \R^n$.
	We then have $n$ coordinate functions 
	$f_1, \ldots, f_n: X \to \R$
	where $f_i = \pi_i \circ f$.
	$f$ is continuous if and only if each of these coordinate functions
	are continuous and, in practise, it is typically easier to check
	these coordinate functions as opposed to checking the entire function
	as a whole.
\end{example}

\begin{proposition}[]
Let $X$ and $Y$ be two non-empty spaces.
$X \times Y$ is Hausdorff if and only if
$X$ is Hausdorff and $Y$ is Hausdorff.
\end{proposition}

\begin{proof}
	$(\impliedby)$: let $(p,q)$ and $(p',q')$ be distinct points of
	$X \times Y$.
	As these points are distinct, $p \neq p'$ or $q \neq q'$.
	If $p \neq p'$, then let open $U, U' \subset X$ such that
	$p \in U$, $p' \in U'$ and $U \cap U' = \varnothing$.
	Then $(p,q) \in U \times Y$ and $(p', q') \in U' \times Y$,
	both of these sets are open and do not intersect.
	This can be done for the case where $q \neq q'$ and we can conclude that
	$X \times Y$ is Hausdorff.

	$(\implies)$: 
	let $p$ and $p'$ be disjoint points of $X$ and $q \in Y$.
	Then there is open $A, A' \subset X \times Y$ with
	$(p,q) \in A$, $(p', q) \in A'$, and $A \cap A' = \varnothing$.
	There is basis elements 
	$U \times V, U' \times V' \in \mathcal B_{X \times Y}$ such that
	$(p,q) \in U \times V \subset A$ and
	$(p',q) \in U' \times V' \subset A'$.
	Observe that $(U \times V) \cap (U' \times V') = \varnothing$.
	But
	$U \times \left\{ q \right\} \subset U \times V$
	and likewise $U' \times \left\{ q \right\} \subset U' \times V' \subset A'$.
	Hence
	\[
		\left( U \times \left\{ q \right\} \right) \cap
		\left( U' \times \left\{ q \right\} \right)
		= \varnothing
	\]
	and so $U \cap U' = \varnothing$.
	As $p \in U$ and $p' \in U'$, $X$ is Hausdorff.
	By symmetry, $Y$ is Hausdorff.
\end{proof}
