%! TEX root = master.tex

\lecture{7}{?}

\begin{definition}[]
	Let $X$ and $Y$ be topological spaces.
	A function $f: X \to Y$ is called \emph{continuous} if for every open
	$U \subset Y$, $f^{-1}(U)$ is open in $X$.
	Let $x \in X$. $f$ is called \emph{continuous at $x$} if for all open
	$U \subset Y$ with $f(x) \in U$, there is open $V \subset X$ with $x \in V$
	such that $f(V) \subset U$.
\end{definition}

\begin{lemma}[]
	Let $X$ and $Y$ be topological space.
	A function $f: X \to Y$ is continuous if and only if for all $x \in X$,
	$f$ is continuous at $x$.
\end{lemma}

\begin{proof}
	Let $f$ be continuous and $x \in X$.
	Let $U \subset Y$ be an open set such that $f(x) \in U$.
	Then $V = f^{-1}(U)$ is also open.
	So $f$ is continuous at $x$.
	Now suppose that for all $x \in X$, $f$ is continuous at $x$.
	Let $U \subset Y$ be open.
	Consider $a \in f^{-1}(U)$.
	$f$ is continuous at $a$ and $f(a) \in U$.
	Hence there is open $V_a \subset f^{-1}(U)$ such that 
	$a \in V_a \subset f^{-1}(U)$.
	We take
	\[
		V = \bigcup_{a \in f^{-1}(U)} V_a.
	\]
	$V$ is open and $f^{-1}(U) = V$.
	Hence $f$ is continuous.
\end{proof}

\begin{example}[]
	Let $X$ be a topological space.
	Consider $A \subset X$ with the subspace topology.
	Then we claim that the \emph{inclusion map} $\iota: A \to X$,
	$\iota(x) = x$, is continuous.
	Proof: let $U \subset X$ be open.
	Then we see that $\iota^{-1}(U) = U \cap A$, which is the intersection of 
	two sets that are open in $A$; hence, it is open.
\end{example}

\begin{lemma}[]
	The composition of continuous functions is a continuous function.
\end{lemma}

\begin{lemma}[]
	Let $X$ and $Y$ be topological spaces and $f: X \to Y$ be continuous.
	Then for all $A \subset Y$ closed, $f^{-1}(A)$ is closed.
\end{lemma}

\begin{proof}
	$\implies$: suppose $A \subset Y$ is closed.
	Then $f^{-1}(A) = X \setminus f^{-1}(Y \setminus A)$.
	As $Y \setminus A$ is open, $f^{-1}(Y \setminus A$ is open.
	So $f^{-1}(A)$ is closed.
	$\impliedby$: Let $U \subset Y$ be open.
	Then $Y \setminus U$ is closed.
	Observe that $f^{-1}(U) = X \setminus f^{-1}(Y \setminus U)$ and 
	$f^{-1}(Y \setminus U)$ is closed and hence $f^{-1}(U)$ is open.
\end{proof}

