\pc{1}{14/11}

\question What is the difference between a host and an end system? List several different types of end systems. Is a web server an end system?

\begin{solution}
    There is no difference between a host and an end system. An end system is just a computer that is connected to a network, so examples include printers, office PCs, personal laptops, etc. A web server is an end system; it has an IP and MAC address just like any other.
\end{solution}


\question Describe the protocol that might be used by two people having a telephone conversation to initiate and end a conversation.

\begin{solution}
   ... 
\end{solution}


\question Is HFC transfer rate dedicated or shared among users? Are collisions possible in a downstream HFC channel? Why or why not?

\begin{solution}
    HFC transfer rate is shared among users through to the cable headend. So there is no dedicated access to the central office. As all packets come from a single source, there is no collisions.   
\end{solution}


\question Describe the different wireless technologies you use during the day and their characteristics. If you have a choice between multiple technologies, why do you prefer one over another?

\begin{solution}
    There are two main wireless technologies in use:
    \begin{enumerate}
        \item WiFi, short distance but high speed connectivity; and
        \item cellular (3G/4G/5G), large distance but medium speed connectivity.
    \end{enumerate}
\end{solution}


\question What are some of the physical media that Ethernet can run over?

\begin{solution}
    Ethernet most commonly runs over twisted-pair copper wire. It also can run over fibers optic links.
\end{solution}


