\documentclass[a4paper, answers]{exam}
\input{../preamble.tex}

\begin{document}

\begin{center}
	\textbf{\textsc{Computational Complexity, Exercises 2}}
\end{center}

\vspace{1em}

\begin{questions}
	\question Prove that \textsc{Dominating Set} is $\mathsf{NP}$-complete.

	Hint for a reduction from \textsc{Vertex Cover}: given a graph $G = (V,E)$,
	create a new graph $H$ as follows: for each edge $(u,v)$ in $G$ add a new
	vertex and join it by an edge to $u$ and $v$.

	\question Prove that \textsc{Dominating Set}, parameterized by $k$, is
	$\textsf{W}[1]$-hard.

	Hint for a reduction from \textsc{Independent Set}: first prove that there is
	a parameterized reduction from \textsc{Independent Set} to
	\textsc{Multicoloured Independent Set} defined as follows: given a graph $G$
	in which every vertex has one of $k$ possible colours (which is not
	necessarily a proper $k$-colouring), does $G$ have an independent set
	containing vertices of all $k$ colours?

	Then use the following construction to provide a parameterized reduction from
	\textsc{Multicoloured Independent Set} to \textsc{Dominating Set}.
	Given a graph $G = (V_G, E_G)$ and a partition 
	$V_G = V_1 \cup \ldots \cup V_k$ (where $V_i$ is the set of vertices coloured
	$i$), create a new graph $H = (V_H, E_H)$ as follows:
	\begin{enumerate}
		\item for every vertex $v \in V_G$, add $v$ to $V_H$;
		\item for each $1 \leq i \leq k$, make $V_i$ a clique in $H$ by adding
			all edges;
		\item for each $1 \leq i \leq k$, introduce two vertices $x_i$ and $y_i$
			into $H$ and make them adjacent to every vertex in $V_i$; and
		\item for every edge $e \in E_G$ with endpoints $u \in V_i$ and 
			$v \in V_j$, introduce a new vertex $w_e$ into $H$ and make it
			adjacent, in $H$, to every vertex in 
			$(V_i \cup V_j) \setminus \left\{ u,v \right\}$.
	\end{enumerate}
	Relate multicoloured independent sets in $G$ to dominating sets in $H$.
	
	Finally, check that a composition of parameterized reductinos is a
	parameterized reduction.

	\question Prove that \textsc{Dominating Set} restricted to graphs of maximum
	degree $6$ (which can also be shown to be $\mathsf{NP}$-complete) is 
	fixed-parameter tractable.

	\question Find a zero knowledge proof protocol based on \textsc{Dominating
	Set} directly, i.e not via reduction from another problem.
	Argue (not just claim) that your protocol has the necessary properties.

	\question Pick your favourite $\mathsf{NP}$-complete problem and find a zero
	knowledge proof based on it (directly).
\end{questions}
\end{document}
