\documentclass[a4paper, answers]{exam}
\input{../preamble.tex}

\begin{document}

\begin{center}
	\textbf{\textsc{Computational Complexity III, Exercises 3}} \\
	\textsc{Ben Napier}
	\vspace{1em}
\end{center}

\begin{questions}
	\question 
	Let $\mathcal L$ be the language of propertly nested brackets.
	For example, $((()))$ and $(()()()(()))$ belong to $\mathcal L$
	but $(()))$ and $)($ do not.
	Show that $\mathcal L \in \mathsf L$.

	\question
	Assuming that the strongest known results about strict inclusion 
	amonst the complexity classes $\mathsf L$, $\mathsf{NL}$,
	$\mathsf{P}$, $\mathsf{NP}$, and $\mathsf{PSPACE}$
	are the Space Hierarchy Theorem and its implications,
	state whether the following statements are true, false, or open
	(in the work you hand in you need only give one word answers).
	\begin{parts}
		\part 
		The language $\mathcal L$ defined in the first question is
		$\mathsf{PSPACE}$-complete.

		\part 
		If 
		$\mathsf P = \mathsf{NP}$ 
		then 
		$\mathsf{NP} \neq \mathsf{PSPACE}$.

		\part
		\textsc{Reachability} is in $\mathsf{PSPACE}$.

		\part 
		If $\mathsf{NL} = \mathsf{P}$,
		then \textsc{Horn-Sat} is not $\mathsf{PSPACE}$-complete.
	\end{parts}

	\question
	Find and explain the error in the following fallacious \emph{proof}
	that $\mathsf P \neq \mathsf{NP}$.
	\begin{proof}
		Assume that $\mathsf P = \mathsf{NP}$.
		Then $\text{SAT} \in \mathsf P$.
		So, for some $k$, $\text{SAT} \in \operatorname{TIME}[n^k]$.
		Because every language in $\mathsf{NP}$ is polynomial time reducible
		to SAT, $\mathsf{NP} \subset \operatorname{TIME}[n^k]$.
		Therefore $\mathsf P \subset \operatorname{TIME}[n^k]$.
		But, by the Time Hierarchy Theorem,
		$\operatorname{TIME}[n^{k+1}]$ contains a language that isn't in
		$\operatorname{TIME}[n^k]$, which contradicts 
		$\mathsf P \subset \operatorname{TIME}[n^k]$.
		Therefore $\mathsf P \neq \mathsf P$.
	\end{proof}

	\question
	Prove that the following problem is $\mathsf{NL}$-complete.
	
	\textsc{Acyclicity} \\
	Instance: a diagraph $G = (V,E)$. \\
	Question: is $G$ acyclic?

	\question
	A $2$-CNF formula $f$ (that is, an instance of $2$-SAT) with variables
	$x_1, \ldots, x_n$ is called \emph{equipartite} if the variables can be
	divided into two sets $A$ and $B$ (each variable belongs to exactly one of
	the sets) such that each clause contains one variable from each of $A$ and
	$B$.
	Prove that the following problem \textsc{Equipartite $2$-Sat} is
	$\mathsf{NL}$-complete.

	\textsc{Equipartite 2-Sat} \\
	Instance: an equipartite $2$-CNF formula $f$. \\
	Question: is $f$ satisfiable?
\end{questions}

\end{document}
