%! TEX root = main.tex

Bioinformatics is the use of computatinoal tools on biological data.
We can catergorise bioinformatics into four main aspects, as follows.
\begin{enumerate}
	\item What is the task?
	\item How do we write a fast algorithm to solve the task?
	\item How can we make our models learn?
	\item How can we differentiate true phenomena from artifacts
		caused by noise?
\end{enumerate}

We will not look at the definition of the task,
this we will leave to the biologists.

\begin{example}[DNA alignment]
	Suppose that we have two strings from the same alphabet.
	We want a metric for how \emph{similar} these two strings are.
\end{example}


Finding the similarity between sequences has significant biological meaning,
it allows us to:
\begin{enumerate}
	\item find similar proteins;
	\item locate similar subsequences; and
	\item locate sequences that may have overlap.
\end{enumerate}

\begin{problem}
	Given $a, b \in \Sigma^\star$,
	what is their optimal alignment?
\end{problem}

\begin{example}[]
	Suppose $a = \mathtt{GCGCA}$ and $b = \mathtt{TGCGC}$.
	One possible alignment is as follows.
	\begin{center}
		\ttfamily
		\begin{tabular}{ccccccc}
			- & G & C & G & C & - & A \\
			T & G & C & G & C & - & -
		\end{tabular}
	\end{center}
\end{example}

But an \emph{optimal} alignment depends on our measure for how good an alignment
is. 
Some possible alignments:
\begin{tasks}(2)
	\task number of perfect matches;
	\task number of mismatches; and
	\task number of mutations.
\end{tasks}
In fact, we can combine all of these measures to create \emph{scoring rules}.
An example:
\begin{tasks}(3)
	\task perfect match: $+1$;
	\task mismatch: $-1$; and
	\task mutation: $-2$.
\end{tasks}
These scoring rules are arbitrary, and in reality the biology behind this
is much more complex (differnt types of mutations can be more or less likely
then others).
If we were to implement a more complex scoring system, we may define a
\emph{scoring function} as
$\sigma: \left( \Sigma \times \left\{ - \right\} \right)^2 \to \R$
where
\begin{enumerate}
	\item $\sigma(x,y)$ is the score of replacing $x$ by $y$;
	\item $\sigma(x,-)$ is the score of deleting $x$; and
	\item $\sigma(-, x)$ is the score of inserting $x$.
\end{enumerate}
Let $A_{ab} = ((a_1,b_1), \ldots, (a_n,b_n))$ be a 
given alignment of two strings $(a_1, \ldots, a_r)$
and $(b_1, \ldots, b_s)$.
Then the \emph{score} of this alignment
is
\[
	\operatorname{score}(A_{ab})
	= \sum_{i=1}^n \sigma(a_i,b_i).
\]
We want to find the \emph{optimal score} of this
alignment, that is, compute
\[
	d(a,b) = \max \operatorname{score}(A_{ab})
\]
An alternative scoring system for a given alignment is \emph{edit distance}.
That is, the minimal cost of edit operations needed to transform one sequence
into another.
An example:
\begin{enumerate}
	\item no change operation: 0;
	\item replace operation: 1; and
	\item add space or delete: 2.
\end{enumerate}
Our goal with this system is to minimise the cost of operations

The total number of alignments for two sequences quickly become very large, so it
is not something we can reasonably brute-force.

\begin{algorithm}[Recursive algorithm for sequence alignment]
	Let $s = (s_1, \ldots, s_{m+1})$, $t = (t_1, \ldots, t_{n+1})$ and $d$ be the
	optimal score of their optimal alignment.
	Note that the last column in the optimal alignment must be one of:
	\begin{enumerate}
		\item $(s_{m+1}, t_{n+1})$, in which case we have
			\[
				d(s, t) = d((s_1,\ldots,s_m), (t_1, \ldots, t_n))
					+ \sigma(s_{m+1}, t_{n+1});
			\]
		\item $(s_{m+1}, -)$, in which case we have
			\[
				d(s,t) = d((s_1, \ldots, s_m), t) + \sigma(s_{m+1},-);
				\;\text{and}
			\]
		\item $(-, t_{n+1})$
			\[
				d(s,t) = d(s, (t_1,\ldots,t_n)) + \sigma(-, t_{n+1}).
			\]
	\end{enumerate}
	We then define the notation
	\[
		V(i,j) = d((s_1, \ldots, s_i), (t_1, \ldots, t_j))
	\]
	and we can use our recursive formula to get
	\begin{align*}
		V(i+1, j+1) &=
		\max\{
			V(i,j) + \sigma(s_{i+1}, t_{j+1}),
			V(i,j+1) + \sigma(s_{i+1},-), \\
			&\qquad V(i+1,j) + \sigma(-,t_{j+1})
		\}.
	\end{align*}
	Our base cases are: $V(0,0) = 0$, $V(i+1,0) = V(i,0) + \sigma(s_{i+1}, -)$,
	and $V(0, j+1) = V(0,j) + \sigma(-, t_{j+1})$.
\end{algorithm}

\begin{example}[]
	Consider the seqeuences $\mathtt{AGC}$ and $\mathtt{AAAC}$.
	Applying the recursive algorithm above, we get the table below.
	\begin{center}
		\ttfamily
		\begin{tabular}{ccccc}
			& & A & G & C \\
			& 0 & -2 & -4 & -6 \\
			A & -2 & 1 & -1 & -3 \\
			A & -4 & -1 & 0 & -2 \\
			A & -6 & -3 & -2 & -1 \\
			C & -8 & -5 & -4 & -1
		\end{tabular}
	\end{center}
	To get the optimal score as $-1$.
	To find the actual alignment, we must backtrack through the maximums we took.
\end{example}

