\documentclass[a4paper, answers]{exam}
\input{../preamble.tex}

\begin{document}

\begin{center}
	\textbf{\textsc{Topology III, Assignment 6}}
	\\
	\textsc{Ben Napier}
	\vspace{1em}
\end{center}

\begin{questions}
	\question
	Suppose that $X$ is a connected topological space
	containing more than one point.
	Suppose further that $\left\{
		p
	\right\}$ is closed for all $p \in X$.
	Show that $X$ has an infinite number of points.
	\begin{solution}
		Assume $X$ has a finite number of points; that is,
		$X = \left\{
			x_1, \ldots, x_n
		\right\}$.
		Let $i \in \left\{
			1, \ldots, n
		\right\}$.
		As $X \setminus \left\{
			x_i
		\right\}$ is open, 
		\[
			\left\{
				x_i
			\right\} = \bigcap_{j \in \left\{
				1, \ldots, n
			\right\} \setminus \left\{
				i
			\right\}} X \setminus \left\{
				x_j
			\right\}
		\]
		is also open as it is a finite intersection.
		Hence, each singleton is open and closed.
		That is, \emph{clopen}.
		But $X$ is connected, so the only sets that are clopen are
		$X$ and $\varnothing$.
		As $X$ contains more than one point, we have a contradiction.
		Thus $X$ has a infinite number of points.
	\end{solution}

	\question
	Suppose $X$ is a connected topological space,
	and suppose that $p \in X$ is such that
	$X \setminus \left\{
		p
	\right\}$
	is disconnected.
	Show that the set $\left\{
		p
	\right\}$ is either open or closed but not clopen.

	Give an example of a connected space $X$ where
	$X \setminus \left\{
		p
	\right\}$
	is disconnected and $\left\{
		p
	\right\}$
	is open, and an example of a connected $X$ where
	$X \setminus \left\{
		p
	\right\}$
	is disconnected and $\left\{
		p
	\right\}$
	is closed.
	\begin{solution}
		First, suppose $\left\{
			p
		\right\}$ is clopen.
		Then $X = \left\{
			p
		\right\} \cup \left( 
			X \setminus \left\{
				p
			\right\} 
		\right)$;
		a contradiction as $X$ is connected.
		Thus $p$ cannot be both open and closed.
		Now, as $X \setminus \left\{
			p
		\right\}$ is disconnected there exists sets $U$ and $V$ which
		are open in $X \setminus \left\{
			p
		\right\}$ such that $U \cup V = X \setminus \left\{
			p
		\right\}$ and $U \cap V = \varnothing$.
		Thus there exists sets $U'$ and $V'$ which are open in $X$ such that
		\[
			U' \cap (X \setminus \left\{
				p
			\right\}) = U' \setminus \left\{
				p
			\right\} = U, \qquad
			V' \cap \left( 
				X \setminus \left\{
					p
				\right\} 
			\right) = V' \setminus \left\{
				p
			\right\} = V.
		\]
		Suppose $U' \cap V' \neq \varnothing$, then we must have
		$U' = U \cup \left\{
			p
		\right\}$ and $V' = V \cup \left\{
			p
		\right\}$.
		Then $U' \cap V' = \left\{
			p
		\right\}$ is open.
		Now suppose $U' \cap V' = \varnothing$.
		As $X$ is connected, $U' \cup V' \neq X$.
		So $U' = U$ and $V' = V$.
		Thus $U' \cup V' = X \setminus \left\{
			p
		\right\}$ is open and so $\left\{
			p
		\right\}$ is closed.

		Consider the set $X = \left\{
			1,2,3
		\right\}$
		with the topology
		\[
			\tau_X = \left\{
				\varnothing,
				X,
				\left\{
					1
				\right\},
				\left\{
					1,2
				\right\},
				\left\{
					1,3
				\right\}
			\right\}
		\]
		and $p = 1$.
		The only clopen sets in $X$ are $\varnothing$ and $X$.
		$X \setminus \left\{
			1
		\right\}$ describes a 2-point set with the discrete topology;
		therefore, it is disconnected.

		Now consider the set $Y = (a,b)$ with the standard topology.
		As $Y$ is an interval, it is connected.
		Now let $p \in (a,b)$.
		As $\R \setminus \left\{
			p
		\right\}$ is open, $\left\{
			p
		\right\}$ is closed.
		Observe
		\[
			X \setminus \left\{
				p
			\right\} = (a,p) \cup (p,b)
		\]
		and thus $X \setminus \left\{
			p
		\right\}$ is disconnected.
	\end{solution}
\end{questions}

\end{document}
