\question Let $H$ be a graph. Prove that the class of $H$-free graphs is closed under edge deletion if and only if $H$ is a complete graph.

\question For $i \geq 0$, the graph $H_i$ is obtained after subdividing one specific edge of the claw $i$ times, so $H_0 = K_{1,3}$, whereas $H_1$ is also known as the chair. Is the class $\mathcal G = \{H_i: i \geq 0\}$ hereditary? If so, determine its set of minimal forbidden induced subgraphs $\mathcal F_{\mathcal G}$.

\question Let $C_5$ denote the $5$-vertex cycle. Determine if the class of $C_5$-free graphs is closed under edge contractions.

\question Let $C_3$ denote the $3$-vertex cycle. Robertson and Seymour gave a cubic-time algorithm for deciding if a graph contains some fixed graph $H$ as a minor. Give a polynomial-time algorithm for deciding if a given graph contains $C_3$ as a minor that does not rely on this result.

\question Determine the clique-width of the graph $G = 2P_5$, that is, $G$ is the disjoint union of two $5$-vertex paths.

\question Let $G$ be a graph and $S$ be a subset of $V(G)$. Let $G - S$ be the graph obtained from $G$ after removing the vertices of $S$. Recall that $S$ is a feedback vertex set if $G - S$ is a forest. The \textsc{Feedback Vertex Set} problem is to determine the size of a smallest feedback vertex set of a given graph. Give a polynomial-time algorithm for \textsc{Feedback Vertex Set} for $10P_1$-free graphs.

\question A graph is co-bipartite graph if it is the complement of a bipartite graph, or in other words, if its vertex set can be partitioned into two (possibly empty) sets $A$ and $B$, such that $A$ and $B$ are cliques. Prove that co-bipartite graphs have unbounded clique-width.

\question Let $P$ be the path on vertices $u_1, \ldots, u_5$ in that order. We obtain a graph $G$ from $P$ by adding four new vertices $s_1, s_2, t_1, t_2$ with edges $u_1s_1, u_1s_2, u_5t_1, u_5t_2$. Determine the treewidth and the pathwidth of $G$. 

\question Let $G$ be a $K_3$-free chordal graph. Give a constant-time algorithm for deciding if $G$ has a $5$-colouring.

\question A graph is a split graph if its vertex set can be partitioned into two (possible empty) sets $C$ and $I$ such that $C$ is a clique and $I$ is an independent set. Express the property of being a split graph in $\operatorname{MSO}_2$.

\question As the property of being split can be expressed in $\operatorname{MSO}_1$, we can decide in polynomial time if a graph from a class of bounded clique-width is a split graph (by applying the meta-theorem on slide 5 of Lecture 19). However, we can recognize \emph{any} split graph in polynomial time. Prove this. 