\chapter{Introduction}
\lecture{1}{7/10}

\begin{definition}[Computer security]
    \textbf{Computer security} is the protection of computer systems against conflicting / attacking / competing environments. 
\end{definition}

\begin{definition}[Asset]
    An \textbf{asset} is something (hardware / software / time / anything) that is of value to an organisation. 
\end{definition}

\begin{definition}[Vulnerability]
    A \textbf{vulnerability} is a weakness of a system that could be exploited to damage assets.
\end{definition}

Vulnerrabilities can arise from many things, such as weak firewall configurations.

\begin{definition}[Threat]
    A \textbf{threat} is the potential danger of someone / something exploiting a vulnerability.
\end{definition}

\begin{definition}[Risk]
    a \textbf{risk} is a calculation (that can be calculated in many differnt ways) which typically takes into account the value of an asset, the threat, and the vulnerability.
\end{definition}

\begin{definition}[Adversaries]
    An \textbf{adversary} is an agent (person, government, etc.) that circumvents the security of a system.
\end{definition}

\begin{definition}[Attack]
    An \textbf{attack} is an \emph{assault} on a system's security.
\end{definition}

We can develop \textbf{attack trees} which are organised plans of attack on a system's security.

\begin{definition}[Countermeasure]
   A \textbf{countermeasure} is an action or process which is used to minimise the risk of a vulnerability. 
\end{definition}

For example, we can use a UPS (uninterrupted power supply) to prevent power cuts having an effect on a system

\begin{definition}[Confidentiality]
    \textbf{Confidentiality} is about ensuring assets are only available to the people it is intended for.
\end{definition}

\begin{definition}[Integrity]
   \textbf{Integrity} is about ensuring the consistency, accurancy, and trustworthness of data. See foreign keys. 
\end{definition}

\begin{definition}[Availability]
    \textbf{Availability} is about ensuring that assets are always avilable.
\end{definition}

\begin{definition}[Accountability]
    \textbf{Accountability} is about ensuring that actions are logged so people can be kept accountable for actions
\end{definition}

\begin{definition}[Reliability]
   \textbf{Reliability} is about ensuring that systems continue running despite errors. 
\end{definition}

