\assignment{4}{14/11}

\setcounter{question}{19}
\question Find the equations of the tangent planes in $\R^3$ to the following surfaces at the given points.
\begin{parts}
    \setcounter{partno}{0}
    \part $x^2 + xy^2 + y^3 + z + 1 = 0$ at $(2, -3, 4)$,
    \begin{solution}
        We set $f(x, y, z) = x^2 + xy^2 + y^3$ and then our surface is defined by the level set $f = -1$. Then, we calculate our grad.
        \[ \nabla f = 
            \bm e_1 (2x + y^2) + 
            \bm e_2 (2xy + 3y^2) + 
            \bm e_3 (1).
        \]
        So $\nabla f(2, -3, 4) = 13\bm e_1 + 15\bm e_2 + \bm e_3$.
        Our tangent plane is normal to this, so it is of the form
        \[ \bm x \cdot \nabla f(2, -3, 4) = d \]
        which, by substituting our point in to find $d$ leaves us with a tangent plane defined by
        \[ 13x + 15y + 1 + 15 = 0. \]
    \end{solution}

    \setcounter{partno}{2}
    \part $\sin(xy) + \sin(yz) + \sin(zx) = 1$ at $(1, \frac\pi2, 0)$.
    \begin{solution}
        We define 
        \[f(x, y, z) = \sin(xy) + \sin(yz) + \sin(zx)\]
        such that our surface is described by the level set $f = 1$. 
        Now
        \[ 
            \nabla f = 
            \bm e_1 (y \cos(xy) + z\cos(zx)) + 
            \bm e_2 (x \cos(xy) + z\cos(yz)) +
            \bm e_3 (y \cos(yz) + x\cos(zx)).
        \]
        Then $\nabla f(1, \frac\pi2, 0) = \bm e_3(\sfrac\pi2 + 1)$.
        Hence, our tangent plane is normal to 
    \end{solution}
\end{parts}

\setcounter{question}{21}
\question What is the derivative of the scalar function $\phi(x, y, z) = x\cos{z} - y$ in the direction $\bm V = \bm e_1 + \bm e_2 + \bm e_3$? 
What is the gradient at the point $(x, y, z) = (0, 1, \sfrac\pi2)$? 
In which direction is $\phi$ increasing the most when moving away from this point?
\begin{solution}
    \begin{align*}
        \nabla \phi &= 
            \bm e_1 (\cos z) - 
            \bm e_2 -
            \bm e_3 (x\sin z). \\
        \norm{V} &= \sqrt 3.
    \end{align*}
    Hence, the directional derivative of $\phi$ in the direction $\bm V$ is
    \[ \norm{V}^{-1} \bm V \cdot \nabla \phi = \frac1{\sqrt 3}(\cos z - 1 - x\sin z). \]
    $\phi(0, 1, \sfrac\pi2) = -\bm e_2$.
    $\phi$ increases the most when moving in the direction $\nabla\phi$, so $-\bm e_2$.
\end{solution}

\setcounter{question}{25}
\question The vector $\bm a$ has components $(a_r) = (1, 1, 1)$ and the vector $\bm b$ has components $(b_r) = (2, 3, 4)$. 
In the following expressions state which indices are free and which are dummy, and give the numerical values of the expressions for each value that the free variable takes (e.g. for $a_r - b_r$ the free variable is $r$ and it takes the value $1$, $2$, $3$ so $a_1 - b_1 = -1$, $a_2 - b_2 = -2$, $a_3 - b_3 = -3$).
\begin{parts}
    \part $a_r + b_r$;
    \begin{solution}
        $r$ is a free variable here. $a_1 + b_1 = 3$, $a_2 + b_2 = 4$, and $a_3 + b_3 = 5$.
    \end{solution}

    \part $a_rb_r$;
    \begin{solution}
        $r$ is a dummy variable here. $a_rb_r = 9$.
    \end{solution}

    \part $a_r b_s a_r$; and
    \begin{solution}
        $r$ is a dummy variable, $s$ is free. $a_rb_1a_r = 6$, $a_rb_2a_r = 9$, and $a_rb_2a_r = 12$.
    \end{solution}

    \part $a_r b_s a_r b_s - a_r b_r a_s b_s$.
    \begin{solution}
        $r$ and $s$ are both dummy variables. $a_rb_sa_rb_s - a_rb_ra_sb_s = -4$.
    \end{solution}
\end{parts}
